Learning constraints in spreadsheets and tabular data

### 2016-09-16 Paper/ECAI/tabular.tex
Various edits

\author{} % \institute{----------------------} }

@@

% why learn them
Such issues and mistakes could be overcome through the development of intelligent user-assisting tools. A core aspect of such tools would be the ability to automatically discover what constraints and formulae are present in a spreadsheet. This is the problem we study in this paper. We discuss multiple promising applications that can build on this, such as rediscovery constraints when importing raw data, auto-suggestion of formula's during spreadsheet use, automatic auto-completion of rows or columns, and consistency and error checking. Our techniques build on work on program synthesis, in particular, on Flashfill \cite{flashfill}, where the definition of a function (over textual cells only) is learned in spreadsheet data from very few examples.

@@

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). Unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and constraints on columns or rows in the sheets.
Furthermore, it derives a set of simple ``atomic'' constraints rather than a set of rules that each consist of a conjunction of literals as in clausal discovery \cite{claudien,lallouet}. Nevertheless, several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy. It also connects with the logical generalization of given formulas in a spreadsheet \cite{Isakowitz}, where one abstracts present formulas by introducing parameters in the place of particular cell references. The key difference here that we start not from existing formulas but from the data directly.

@@

  \caption{Constraints learned for the example (left), except for \textit{ALLDIFFERENT} (18), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5).

@@

  \item \CFunction: each value in the vector $\sbl{r}$ is obtained by summing over the corresponding row in $\bsbl{x}$.
  %For example, $\sbl{r} = [12, 7]$ and $\bsbl{x}$ consists of columns: $[5, 15]$, $[7, -8]$.
\end{itemize}
\textit{SUM} is an aggregate constraint, and a similar constraint is available for aggregate operators \textit{MIN}, \textit{MAX}, \textit{AVERAGE}, \textit{PRODUCT} and \textit{COUNT}. The \textit{conditional} aggregate variant \textit{SUMIF} only ranges over those cells that satisfy a condition on a related cell in that table, e.g. $\range{T_1}{\rangeall}{10}$ and the data in $T_5$ for matching salesperson names.
\end{example}

@@

\begin{table}[t]
  \caption{Translation of signature requirements to superblock constraints. The following need not be relaxed: $\plength(B) = x$, $\por(B) = x$, $\ptable(B) = x$ and $\psize(B) \geq x$.}
  \label{tbl:translation}
  \begin{tabularx}{\linewidth}{lX}
    \textbf{Requirement} & \textbf{Superblock constraint} \\ \hline \hline
    $\ptype(B) = x$ & $\mathit{basetype}(\ptype(B)) = \mathit{basetype}(x)$ \\ \hline
    $\psize(B) = x$ & $\psize(B) \geq x$ \\ \hline
    $\pcols(B) = x$ & $\mathit{if}~\por(B) = \mathit{column}: \pcols(B) \geq x$ \\ 
    & $\mathit{if}~\por(B) = \mathit{row}: \pcols(B) = x$ \\ \hline
    $\prows(B) = x$ & $\mathit{if}~\por(B) = \mathit{column}: \prows(B) = x$ \\ 
    & $\mathit{if}~\por(B) = \mathit{row}: \prows(B) \geq x$
  \end{tabularx}
\end{table}

@@

\subsubsection{Generating subblock assignments (Step 3b)}
\label{sec:algo:subgr}
Given a superblock assignment $(\sbl{1}, \dots, \sbl{n})$ from the previous step, the goal is to discover valid \textit{subassignments}, i.e. assignments of subblocks $(\ssbl{1}, \dots, \ssbl{n})$ (for all $i$, $\ssbl{i} \sqsubseteq \sbl{i}$) that satisfy both the signature and the definition of template~$s$.

@@

@@

\begin{algorithm}[t]
  \begin{algorithmic}
    \footnotesize
    \Procedure{Subassignments}{$s$, $(\sbl{1}, \dots, \sbl{n})$}
    \State $n \gets$ number of arguments of template $s$
    \State $A_{sub} \gets \emptyset$
    \ForAll{$(\sbl{1}', \dots, \sbl{n}') \textbf{ where } \forall i: \sbl{i}' \sqsubseteq \sbl{i}  \land \textit{disjoint}_{j \neq i}(\sbl{i}', \sbl{j}')$}
      \If{$\textit{Sig}_s(\sbl{1}',\dots,\sbl{n}') \land \textit{Def}_s(\sbl{1}',\dots,\sbl{n}')$}
        \State $A_{sub} \gets A_{sub} \cup \{ (\sbl{1}', \dots, \sbl{n}') \}$
      \EndIf
    \EndFor
\Return $A_{sub}$
\EndProcedure
\end{algorithmic}
\caption{Generate-and-test for $\findassignment$}
\label{algo:gen-test-subblocks}
\end{algorithm}

@@

For \textit{FOREIGNKEY}, \textit{LOOKUP}, aggregates and \textit{PROJECT} we use custom, optimized methods to find subassignments.
Instead of the more generic generate-and-test implementation, we use, for example, caching, maximal range-finding and partial tests.
While conditional aggregates use generate-and-test, they employ similar optimizations in their test method.

@@

For this example our primary goal of rediscovering all intended constraints is achieved, the recall (i.e. $\frac{\text{intended discovered}}{\text{all intended}}$) is~$1.0$.
The implementation also returns 6 additional constraints, therefore, our precision (i.e. $\frac{\text{intended discovered}}{\text{all discovered}}$) is $12/18 = 0.67$, which is mostly due to the spurious RANK constraints. We believe this is due to the many short vectors which increases the chance for constraints to be true by accident.

@@

\samuel{fail-fast?}
Of the 32 supported constraint templates (including equality), the slowest by far is fuzzy lookup using $\pm 36\%$ of the running time.
It is followed by product ($\pm 5\%$), relative difference ($\pm 5\%$) and various row-aggregates (each $3.5\% - 4\%$).

@@

There are multiple factors that impact the running time of \sname: the number of vectors, the size of the vectors, the number of vectors per block, the type of cells, the number of tables and the number of constraints that are in the data.
Moreover, the running time depends on the number of the constraint templates implemented in the system and their arity.

In this section the impact on the running time is analyzed of 1)~the number of vectors; 2)~the size of the vectors; and 3)~the size of the blocks.
This analysis is supported by a set of experiments we have run on synthetic spreadsheets.
These spreadsheets contains random integers\footnote{For strings and floats the results are similar but the running times are typically much lower.} and are made up of two tables.
For one table the number of vectors ($2$) and block size ($2$) is fixed, while for the other table these properties are varied.
Moreover, the size of vectors (for both tables) can be adapted, while the default is~$8$.
The running times for various configurations are plotted in Figure~\ref{fig:runtime_analysis}.

For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of size~$k$, the worst case complexity becomes $O(n * m * k)$.
However, in practice, most subassignments will not satisfy the constraint template.
Since our implementation tries to account for this by failing fast, the amortized time is constant in the size of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small). 
The benchmark experiments show that the number of cells in a spreadsheet has a very limited impact on the running time.
As shown in Figure~\ref{fig:runtime_analysis}b, the synthetic experiments confirm that increasing the size of the vectors has a negligible impact on the performance for most template.
Most of the increase in running time can be attributed to the RANK constraint which is less suited for failing fast.

On the contrary, the number of vectors in a spreadsheet have a significant impact on the running time.
Many constraints take single vector arguments, therefore, the number of subassignments to be tested depends on the number of vectors in the spreadsheets and the arity of the constraint.
For a constraint template with arity~$n$ and $v$~vectors there are $n^v$ subassignments (or $\frac{n!}{(n - v)!}$ if the arguments have to be different).
As the number of vectors in a spreadsheet increases, the running time will quickly be dominated by constraints with a larger arity.
This effect is noticeable in our synthetic experiments when the number of (equally sized) vectors is increased (Figure~\ref{fig:runtime_analysis}a)
However, in our benchmark experiments there will be various factors that limit the rapid increase in running time because the assignment step can disqualify entire blocks that do not match the signature requirements, i.e. blocks of vectors with different lengths for the product constraint.

@@

These experiments, however, also show that the running time for various non-aggregate constraints increases as the size of blocks becomes very small ($1$ or $2$).
Since these constraints have single vector arguments, the block size should not have a strong effect.
However, in practice, for very small blocks the superblock assignment phase has to generate a very large number of combinations, while for larger blocks this work is done mostly during the subassignment generation phase.
Since the implementation of that last phase is more adept at generating these combinations and does not need to check for constraints of all these combinations, shifting this workload to the assignment phase slows down the algorithm.

>>>>

\author{Samuel Kolb \and Sergey Paramonov \and Thias Guns \and Luc De Raedt} % \institute{KULeuven}

@@

% why learn them
Such issues and mistakes could be overcome through the development of intelligent user-assisting tools. A core aspect of such tools would be the ability to automatically discover what constraints and formulae are present in a spreadsheet. This is the problem we study in this paper. We discuss multiple promising applications that can build on this, such as rediscovery of constraints when importing raw data, auto-suggestion of formula's during spreadsheet use, automatic auto-completion of rows or columns, and consistency and error checking. Our techniques build on work on program synthesis, in particular, on Flashfill \cite{flashfill}, where the definition of a function (over textual cells only) is learned in spreadsheet data from very few examples.

@@

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). Unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and constraints on columns or rows in the sheets.
Furthermore, it derives a set of simple ``atomic'' constraints rather than a set of rules that each consist of a conjunction of literals as in clausal discovery \cite{claudien,lallouet}. Nevertheless, several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy. It also connects with the logical generalization of given formulas in a spreadsheet \cite{Isakowitz}, where one abstracts present formulas by introducing parameters in the place of particular cell references. The key difference here is that we start not from existing formulas but from the data directly.

@@

  \caption{Constraints learned for the example (top), except for \textit{ALLDIFFERENT} (18), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5).

@@

  \item \CFunction: each value in the vector $\sbl{r}$ is obtained by summing over the corresponding row in $\bsbl{x}$.
  %For example, $\sbl{r} = [12, 7]$ and $\bsbl{x}$ consists of columns: $[5, 15]$, $[7, -8]$.
\end{itemize}
\textit{SUM} is an aggregate constraint, and a similar constraint is available for aggregate operators \textit{MIN}, \textit{MAX}, \textit{AVERAGE}, \textit{PRODUCT} and \textit{COUNT}. The \textit{conditional} aggregate variant \textit{SUMIF} only ranges over those cells that satisfy a condition on a related cell in that table, e.g. for the spreadsheet in Figure~\ref{fig:main_example}: $\range{T_1}{\rangeall}{10}$ and the data in $T_5$ for matching salesperson names.
\end{example}

@@

\begin{table}[tbh]
  \caption{Translation of signature requirements to superblock constraints. The following need not be relaxed: $\plength(B) = x$, $\por(B) = x$, $\ptable(B) = x$ and $\psize(B) \geq x$.}
  \label{tbl:translation}
  \begin{tabularx}{\linewidth}{lX}
    \textbf{Requirement} & \textbf{Superblock constraint} \\ \hline \hline
    $\ptype(B) = x$ & $\mathit{basetype}(\ptype(B)) = \mathit{basetype}(x)$ \\ \hline
    $\psize(B) = x$ & $\psize(B) \geq x$ \\ \hline
    $\pcols(B) = x$ & $\mathit{if}~\por(B) = \mathit{column}: \pcols(B) \geq x$ \\ 
    & $\mathit{if}~\por(B) = \mathit{row}: \pcols(B) = x$ \\ \hline
    $\prows(B) = x$ & $\mathit{if}~\por(B) = \mathit{column}: \prows(B) = x$ \\ 
    & $\mathit{if}~\por(B) = \mathit{row}: \prows(B) \geq x$
  \end{tabularx}
\end{table}

@@

\subsubsection{Generating subblock assignments (Step 3b)}
\label{sec:algo:subgr}

Given a superblock assignment $(\sbl{1}, \dots, \sbl{n})$ from the previous step, the goal is to discover valid \textit{subassignments}, i.e. assignments of subblocks $(\ssbl{1}, \dots, \ssbl{n})$ (for all $i$, $\ssbl{i} \sqsubseteq \sbl{i}$) that satisfy both the signature and the definition of template~$s$.

@@

\begin{algorithm}[tbh]
  \begin{algorithmic}
    \footnotesize
    \Procedure{Subassignments}{$s$, $(\sbl{1}, \dots, \sbl{n})$}
    \State $n \gets$ number of arguments of template $s$
    \State $A_{sub} \gets \emptyset$
    \ForAll{$(\sbl{1}', \dots, \sbl{n}') \textbf{ where } \forall i: \sbl{i}' \sqsubseteq \sbl{i}  \land \textit{disjoint}_{j \neq i}(\sbl{i}', \sbl{j}')$}
      \If{$\textit{Sig}_s(\sbl{1}',\dots,\sbl{n}') \land \textit{Def}_s(\sbl{1}',\dots,\sbl{n}')$}
        \State $A_{sub} \gets A_{sub} \cup \{ (\sbl{1}', \dots, \sbl{n}') \}$
      \EndIf
    \EndFor
\Return $A_{sub}$
\EndProcedure
\end{algorithmic}
\caption{Generate-and-test for $\findassignment$}
\label{algo:gen-test-subblocks}
\end{algorithm}

@@

@@

For \textit{FOREIGNKEY}, \textit{LOOKUP}, aggregates and \textit{PROJECT} we use custom, optimized methods to find subassignments.
Instead of the more generic generate-and-test implementation, we use, for example, caching, maximal range-finding and partial tests.
While conditional aggregates use generate-and-test, they employ similar optimizations in their test method.
Based on the assumptions that most subassignments will not satisfy a constraint template, the implementations are geared to fail-fast when possible.

@@

For this example our primary goal of rediscovering all intended constraints is achieved, the recall (i.e. $\frac{\text{intended discovered}}{\text{all intended}}$) is~$1.0$.
The implementation also returns 6 additional constraints, therefore, our precision (i.e. $\frac{\text{intended discovered}}{\text{all discovered}}$) is $12/18 = 0.67$, which is mostly due to the spurious RANK constraints.
This spreadsheet contains many short vectors which increases the chance for constraints to be true by accident.

@@

Of the 32 supported constraint templates (including equality), the slowest by far is fuzzy lookup using $\pm 36\%$ of the running time.
It is followed by product ($\pm 5\%$), relative difference ($\pm 5\%$) and various row-aggregates (each $3.5\% - 4\%$).

@@

There are multiple factors that impact the running time of \sname: the number of vectors, the length of the vectors, the number of vectors per block, the type of cells, the number of tables and the number of constraints that are in the data.
Moreover, the running time depends on the number of the constraint templates implemented in the system and their arity.

In this section the impact on the running time is analyzed of 1)~the number of vectors; 2)~the length of the vectors; and 3)~the size of the blocks.
This analysis is supported by a set of experiments we have run on synthetic spreadsheets.
These spreadsheets contains random integers\footnote{For strings and floats the results are similar but the running times are typically much lower.} and are made up of two tables.
For one table the number of vectors ($2$) and block size ($2$) is fixed, while for the other table these properties are varied.
Moreover, the length of the vectors (for both tables) can be adapted, while the default is~$8$.
The running times for various configurations are plotted in Figure~\ref{fig:runtime_analysis}.

For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of length~$k$, the worst case complexity becomes $O(n * m * k)$.
However, in practice, most subassignments will not satisfy the constraint template.
Since our implementation tries to account for this by failing fast, the amortized time is constant in the length of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small). 
The benchmark experiments show that the number of cells in a spreadsheet has a very limited impact on the running time.
As shown in Figure~\ref{fig:runtime_analysis}b, the synthetic experiments confirm that increasing the length of the vectors has a negligible impact on the performance for most template.
Most of the increase in running time can be attributed to the RANK constraint which is less suited for failing fast.

On the contrary, the number of vectors in a spreadsheet have a significant impact on the running time.
Many constraints take single vector arguments, therefore, the number of subassignments to be tested depends on the number of vectors in the spreadsheets and the arity of the constraint.
For a constraint template with arity~$n$ and $v$~vectors there are $n^v$ subassignments (or $\frac{n!}{(n - v)!}$ if the arguments have to be different).
As the number of vectors in a spreadsheet increases, the running time will quickly be dominated by constraints with a larger arity.
This effect is noticeable in our synthetic experiments when the number of (equally sized) vectors is increased (Figure~\ref{fig:runtime_analysis}a).
In practice, in our benchmark experiments, the assignment step of our algorithm is able to limit this rapid increase in running time for larger numbers of vectors.
The superblock assignment phase is able to disqualify entire blocks that do not match the signature requirements, i.e. blocks of vectors with different lengths for the product constraint.

@@

These experiments, however, also show that the running time for various non-aggregate constraints increases as the size of blocks becomes small.
Since these constraints have single vector arguments, the block size should not have a strong effect.
However, in practice, when there are many small blocks, the superblock assignment phase has to generate a very large number of combinations, while for larger blocks this work is done mostly during the subassignment generation phase.
Since the implementation of that last phase is more adept at generating these combinations and does not need to check for constraints on all these combinations, shifting this workload to the assignment phase slows down the algorithm.
